---
title: "RNAseq local response"
author: "Sara Hoefer"
date: "2026-07-27"
output: html_document
---

# 1. Packages
```{r}
library(pheatmap)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(RColorBrewer) # to change colors in pheatmap
library(tidyverse)

## to install DESeq2:
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("DESeq2")
```

# 2. Functions
```{r}
se <- function (x) {sd (x,na.rm =TRUE) / sqrt (length(na.exclude (x)))}
```


# 3. Data
## 3.1. Load data

```{r}
dat_raw_counts <- read.table("03_local_response_RNAseq/01_data/within_leaf_analysis_ready_gene_counts.tsv",
                             header = T, row.names = 1, sep = "\t")

dat_fpkm <- read.table("03_local_response_RNAseq/01_data/within_leaf_analysis_ready_gene_fpkm_all_samples.tsv",
                             header = T, row.names = 1, sep = "\t")

dat_goi <- read.table("03_local_response_RNAseq/01_data/tristan_heatmap_gene_categories.csv",
                      header = T, sep = ",")

dat_phytohormone_goi <- read.table("03_local_response_RNAseq/01_data/phytohormone_candidates_gene_list.csv",
                                   header = T, sep = ",")

dat_meta <- read.table("03_local_response_RNAseq/01_data/within_leaf_analysis_ready_info.csv",
                             header = T, sep = ",")
```

## 3.2. Wrangle data

```{r}
# create col segment in metadata

dat_meta$segment <- sub(".*(S.).*", "\\1", dat_meta$sample_description)


# change phytohormone Gene_ID_v5 to Gene_ID

dat_ph_goi <- dat_phytohormone_goi |>
  mutate(Gene_ID = Gene_ID_v5)
```

```{r}
# subset by genes of interest

# dat_subset <- dat_raw_counts[rownames(dat_raw_counts) %in% dat_goi$Gene_ID, ]

# dat_subset <- dat_raw_counts[rownames(dat_raw_counts) %in% dat_phytohormone_goi$Gene_ID_v5, ]

```

# 4. Heatmap
## 4.1. DESeq2 for heatmap

```{r}

# 1. Create DESeq2 object (Replace with your actual data frames)
# countData: matrix of raw gene counts
# colData: dataframe with sample metadata, last condition is the default comparison (so it's treat vs. ctrl)

genotypes <- unique(dat_meta$genotype)
segments <- unique(dat_meta$segment)

results_list <- list()
dds_list <- list()

for(g in genotypes){
  for(s in segments){

    keep <- dat_meta$genotype == g &
            dat_meta$segment == s

    dds_sub <- DESeqDataSetFromMatrix(
      countData = dat_subset_ph[, keep],
      colData = dat_meta[keep, ],
      design = ~ condition_group
    )

    dds_sub <- DESeq(dds_sub)

    res <- results(dds_sub,
                   contrast = c("condition_group", "w", "c"))

    results_list[[paste(g, s, sep = "_")]] <- res
    dds_list[[paste(g, s, sep = "_")]] <- dds_sub
  }
}

```

```{r}
# create matrix for heatmap

for (name in c(
  "kn_S1", "kn_S2", "kn_S3",
  "lox_S1", "lox_S2", "lox_S3",
  "haci_S1", "haci_S2", "haci_S3"
)) {

  # DESeq2 results
  res <- as.data.frame(results_list[[name]])

  # VST transform
  vst_data <- varianceStabilizingTransformation(
    dds_list[[name]],
    blind = FALSE
  )

  vst_matrix <- assay(vst_data)

  # scale each gene
  heatmap_matrix <- t(scale(t(vst_matrix)))

  # get treatment labels
  condition <- colData(dds_list[[name]])$condition_group

  # average replicates
  heatmap_avg <- sapply(
    levels(condition),
    function(x) {
      rowMeans(heatmap_matrix[, condition == x, drop = FALSE])
    }
  )

  # reorder rows according to dat_phytohormone_goi order
  row_order <- match(dat_phytohormone_goi$Gene_ID_v5, rownames(heatmap_avg))
  row_order <- row_order[!is.na(row_order)]

  heatmap_avg <- heatmap_avg[row_order, ]

  # replace row names with gene names
  gene_names <- dat_phytohormone_goi$gene_name[
    match(rownames(heatmap_avg), dat_phytohormone_goi$Gene_ID_v5)
  ]

  rownames(heatmap_avg) <- gene_names
  colnames(heatmap_avg) <- c("Control", "MW")


  ## star matrix for this name only
  
  star_matrix <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = 1,
    dimnames = list(
      rownames(heatmap_avg),
      name
    )
  )

  res <- results_list[[name]]

  common <- intersect(rownames(res), rownames(star_matrix))

  padj <- res[common, "padj"]

  star_matrix[common[!is.na(padj) & padj < 0.001], name] <- "***"
  star_matrix[common[!is.na(padj) & padj < 0.01 & padj >= 0.001], name] <- "**"
  star_matrix[common[!is.na(padj) & padj < 0.05 & padj >= 0.01], name] <- "*"


  # make stars match heatmap dimensions
  stars <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = ncol(heatmap_avg),
    dimnames = dimnames(heatmap_avg)
  )

  stars[, "MW"] <- star_matrix[, name]


  ## save heatmap
  
  # pdf(paste0(name, "_hm2.pdf"), width = 5, height = 12)

  pheatmap(
    heatmap_avg,
    display_numbers = stars,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    show_rownames = TRUE,
    fontsize_col = 12,
    fontsize_row = 10,
    cellwidth = 10,
    cellheight = 10,
    main = name
  )

  # dev.off()
}
```

# 5. Barplots log-fold change

## 5.1. Subset of genes

Loop for everything from ChatGPT:

```{r}

# ---------------------------------------------------------
# Prepare metadata
# ---------------------------------------------------------

rownames(dat_meta) <- dat_meta$sample_description

# Make sure count matrix follows metadata order
dat_raw_counts <- dat_raw_counts[, rownames(dat_meta)]

# Factors
dat_meta$genotype <- factor(
  dat_meta$genotype,
  levels = c("kn", "lox", "haci")
)

dat_meta$condition_group <- factor(
  dat_meta$condition_group,
  levels = c("w", "c")
)

dat_meta$segment <- factor(
  dat_meta$segment,
  levels = c("S1", "S2", "S3")
)


# ---------------------------------------------------------
# Function
# ---------------------------------------------------------

make_segment_plot <- function(segment_name,
                              counts,
                              metadata,
                              goi,
                              padj_cutoff = 0.05) {
  
  # -------------------------------------------------------
  # 1. Subset metadata and counts for this segment
  # -------------------------------------------------------
  
  meta_seg <- metadata %>%
    filter(segment == segment_name)
  
  counts_seg <- counts[, rownames(meta_seg)]
  
  
  # -------------------------------------------------------
  # 2. Create DESeq2 object
  # -------------------------------------------------------
  
  dds <- DESeqDataSetFromMatrix(
    countData = round(as.matrix(counts_seg)),
    colData = meta_seg,
    design = ~ genotype + condition_group +
              genotype:condition_group
  )
  
  
  # -------------------------------------------------------
  # 3. Filter low-count genes
  # -------------------------------------------------------
  
  keep <- rowSums(counts(dds) >= 10) >= 2
  dds <- dds[keep, ]
  
  
  # -------------------------------------------------------
  # 4. Run DESeq2
  # -------------------------------------------------------
  
  dds <- DESeq(dds)
  
  
  # -------------------------------------------------------
  # 5. Interaction tests
  #
  # Does the C-vs-W response differ between:
  #   WT vs lox
  #   WT vs haci
  # -------------------------------------------------------
  
  res_lox_interaction <- results(
    dds,
    name = "genotypelox.condition_groupc"
  )
  
  res_haci_interaction <- results(
    dds,
    name = "genotypehaci.condition_groupc"
  )
  
  
  # -------------------------------------------------------
  # 6. Significant interaction genes
  # -------------------------------------------------------
  
  sig_lox <- rownames(res_lox_interaction)[
    !is.na(res_lox_interaction$padj) &
      res_lox_interaction$padj < padj_cutoff
  ]
  
  sig_haci <- rownames(res_haci_interaction)[
    !is.na(res_haci_interaction$padj) &
      res_haci_interaction$padj < padj_cutoff
  ]
  
  
  # Significant in WT-vs-lox OR WT-vs-haci
  sig_genes <- union(
    sig_lox,
    sig_haci
  )
  
  
  # -------------------------------------------------------
  # 7. Restrict to genes of interest
  # -------------------------------------------------------
  
  sig_genes <- intersect(
    sig_genes,
    goi$Gene_ID
  )
  
  
  # -------------------------------------------------------
  # 8. Get C vs W effect for WT
  # -------------------------------------------------------
  
  res_WT <- results(
    dds,
    contrast = c(
      "condition_group",
      "c",
      "w"
    )
  )
  
  
  # -------------------------------------------------------
  # 9. Get C vs W effect for lox
  # -------------------------------------------------------
  
  res_lox <- results(
    dds,
    contrast = list(
      c(
        "condition_group_c_vs_w",
        "genotypelox.condition_groupc"
      )
    )
  )
  
  
  # -------------------------------------------------------
  # 10. Get C vs W effect for haci
  # -------------------------------------------------------
  
  res_haci <- results(
    dds,
    contrast = list(
      c(
        "condition_group_c_vs_w",
        "genotypehaci.condition_groupc"
      )
    )
  )
  
  
  # -------------------------------------------------------
  # 11. Create plotting dataframe
  #
  # Keep both log2FC and lfcSE
  # -------------------------------------------------------
  
  plot_data <- bind_rows(
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "WT",
      log2FC = res_WT[sig_genes, "log2FoldChange"],
      lfcSE = res_WT[sig_genes, "lfcSE"]
    ),
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "lox",
      log2FC = res_lox[sig_genes, "log2FoldChange"],
      lfcSE = res_lox[sig_genes, "lfcSE"]
    ),
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "haci",
      log2FC = res_haci[sig_genes, "log2FoldChange"],
      lfcSE = res_haci[sig_genes, "lfcSE"]
    )
  )
  
  # ------------------------------------------------------- 
  # 12. Add interaction p-values to plotting data
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    padj_lox_vs_WT = res_lox_interaction[Gene_ID, "padj"],
    padj_haci_vs_WT = res_haci_interaction[Gene_ID, "padj"]
  )
  
  
  # -------------------------------------------------------
  # 13. Create a significance label:
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    significance = case_when(
      genotype == "lox" & padj_lox_vs_WT < 0.001  ~ "***",
      genotype == "lox" & padj_lox_vs_WT < 0.01   ~ "**",
      genotype == "lox" & padj_lox_vs_WT < 0.05   ~ "*",
      
      genotype == "haci" & padj_haci_vs_WT < 0.001  ~ "***",
      genotype == "haci" & padj_haci_vs_WT < 0.01   ~ "**",
      genotype == "haci" & padj_haci_vs_WT < 0.05   ~ "*",
      
      TRUE ~ ""
    )
  )
  
  # -------------------------------------------------------
  # 14. Calculate vjust in data:
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    star_y = ifelse(
      log2FC >= 0,
      log2FC + 1.96 * lfcSE,
      log2FC - 1.96 * lfcSE
    ),
    star_vjust = ifelse(
      log2FC >= 0,
      -0.3,
      1.3
    )
  )
  
  
  # -------------------------------------------------------
  # 15. Add gene names
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
    left_join(
      goi %>%
        select(Gene_ID, gene_name),
      by = "Gene_ID"
    )
  
  
  # -------------------------------------------------------
  # 16. Plot
  # -------------------------------------------------------
  
  p <- ggplot(
    plot_data,
    aes(
      x = gene_name,
      y = log2FC,
      fill = genotype
    )
  ) +
    
    geom_col(
      position = position_dodge(width = 0.8)
    ) +
    
    geom_errorbar(
      aes(
        ymin = log2FC - 1.96 * lfcSE,
        ymax = log2FC + 1.96 * lfcSE
      ),
      position = position_dodge(width = 0.8),
      width = 0.25
    ) +
    
    geom_text(
      aes(
        label = significance,
        y = star_y,
        vjust = star_vjust
      ),
      position = position_dodge(width = 0.8),
      size = 4
    ) +
    
    geom_hline(
      yintercept = 0,
      linewidth = 0.5
    ) +
    
    theme_classic() +
    
    labs(
      title = segment_name,
      x = NULL,
      y = "log2 fold change (C vs W)",
      fill = "Genotype"
    ) +
    
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5
      )
    )
  
  
  # -------------------------------------------------------
  # 17. Return everything
  # -------------------------------------------------------
  
  return(
    list(
      dds = dds,
      significant_genes = sig_genes,
      interaction_lox = res_lox_interaction,
      interaction_haci = res_haci_interaction,
      plot_data = plot_data,
      plot = p
    )
  )
}
```


```{r}
# directly change input here

S1 <- make_segment_plot(
  segment_name = "S1",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)

S2 <- make_segment_plot(
  segment_name = "S2",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)

S3 <- make_segment_plot(
  segment_name = "S3",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)
```

```{r}
fig_S1 <- S1$plot
fig_S2 <- S2$plot
fig_S3 <- S3$plot

fig_S1
fig_S2
fig_S3
```

```{r}
# save plots:

# pdf(
#   "S1_barplot2.pdf",
#   width = 12,
#   height = 5
# )
# 
# print(fig_S1)
# 
# dev.off()
```

## 5.2. All genes

Find all genes that have significantly different lfc between mutant and wt, for both mutants, per segment

```{r}

```






# BS:
# Subset of genes that are interesting
## 1. Stats

```{r}

# 1. Create DESeq2 object (Replace with your actual data frames)
# countData: matrix of raw gene counts
# colData: dataframe with sample metadata, last condition is the default comparison (so it's treat vs. ctrl)

genotypes <- unique(dat_meta$genotype)
segments <- unique(dat_meta$segment)

results_list <- list()
dds_list <- list()

for(g in genotypes){
  for(s in segments){

    keep <- dat_meta$genotype == g &
            dat_meta$segment == s

    dds_sub <- DESeqDataSetFromMatrix(
      countData = dat_subset_ph[, keep],
      colData = dat_meta[keep, ],
      design = ~ condition_group
    )

    dds_sub <- DESeq(dds_sub)

    res <- results(dds_sub,
                   contrast = c("condition_group", "w", "c"))

    results_list[[paste(g, s, sep = "_")]] <- res
    dds_list[[paste(g, s, sep = "_")]] <- dds_sub
  }
}

```

```{r}
# names(results_list)
# head(results_list[["lox_S2"]])
# 
# sapply(results_list, nrow)
# 
# View(as.data.frame(results_list[["kn_S1"]]))
```


```{r}
# for(name in names(results_list)){
#   
#   res <- as.data.frame(results_list[[name]])
#   
#   gene_names <- dat_phytohormone_goi$gene_name[
#     match(rownames(res), dat_phytohormone_goi$Gene_ID_v5)
#   ]
#   
#   # keep Gene_ID if no name found
#   gene_names[is.na(gene_names)] <- rownames(res)[is.na(gene_names)]
#   
#   rownames(res) <- gene_names
#   
#   results_list[[name]] <- res
# }
```

star matrix for heatmap:
```{r}

# # star_matrix <- matrix("", nrow = 3, ncol = 61)
# 
# star_matrix[,] <- ""
# 
# for(name in names(results_list)){
#   
#   res <- results_list[[name]]
#   
#   common <- intersect(rownames(res), rownames(star_matrix))
#   
#   padj <- res[common, "padj"]
#   
#   # *** padj < 0.001
#   sig_genes <- common[!is.na(padj) & padj < 0.001]
#   star_matrix[sig_genes, name] <- "***"
#   
#   # ** padj < 0.01
#   sig_genes <- common[!is.na(padj) & padj < 0.01 & padj >= 0.001]
#   star_matrix[sig_genes, name] <- "**"
#   
#   # * padj < 0.05
#   sig_genes <- common[!is.na(padj) & padj < 0.05 & padj >= 0.01]
#   star_matrix[sig_genes, name] <- "*"
# }
# 
# # NOW convert Gene_ID rownames to gene_name
# gene_names <- dat_goi$gene_name[match(rownames(star_matrix),
#                                       dat_goi$Gene_ID)]
# 
# gene_names[is.na(gene_names)] <- rownames(star_matrix)[is.na(gene_names)]
# 
# rownames(star_matrix) <- gene_names
# rownames(heatmap_avg) <- gene_names
```

```{r}
# For barplots:
## Check which log2 fold change are significantly different from KN in mutants
segments <- unique(dat_meta$segment)

dds_list <- list()
resultsNames_list <- list()

for(s in segments){

  keep <- dat_meta$segment == s

  dds <- DESeqDataSetFromMatrix(
    countData = dat_subset_ph[, keep],
    colData = dat_meta[keep, ],
    design = ~ genotype * condition_group
  )

  dds$genotype <- relevel(factor(dds$genotype), ref = "kn") 
  dds$condition_group <- relevel(factor(dds$condition_group), ref = "c")  # control reference

  dds <- DESeq(dds)

  dds_list[[s]] <- dds
  resultsNames_list[[s]] <- resultsNames(dds)
}
```

```{r}
# resultsNames(dds_list[["leaf"]])
```

## 2. Heatmap

Loop for all 9 at once:

```{r}
for (name in c(
  "kn_S1", "kn_S2", "kn_S3",
  "lox_S1", "lox_S2", "lox_S3",
  "haci_S1", "haci_S2", "haci_S3"
)) {

  # DESeq2 results
  res <- as.data.frame(results_list[[name]])

  # VST transform
  vst_data <- varianceStabilizingTransformation(
    dds_list[[name]],
    blind = FALSE
  )

  vst_matrix <- assay(vst_data)

  # scale each gene
  heatmap_matrix <- t(scale(t(vst_matrix)))

  # get treatment labels
  condition <- colData(dds_list[[name]])$condition_group

  # average replicates
  heatmap_avg <- sapply(
    levels(condition),
    function(x) {
      rowMeans(heatmap_matrix[, condition == x, drop = FALSE])
    }
  )

  # reorder rows according to dat_phytohormone_goi order
  row_order <- match(dat_phytohormone_goi$Gene_ID_v5, rownames(heatmap_avg))
  row_order <- row_order[!is.na(row_order)]

  heatmap_avg <- heatmap_avg[row_order, ]

  # replace row names with gene names
  gene_names <- dat_phytohormone_goi$gene_name[
    match(rownames(heatmap_avg), dat_phytohormone_goi$Gene_ID_v5)
  ]

  rownames(heatmap_avg) <- gene_names
  colnames(heatmap_avg) <- c("Control", "MW")


  ## star matrix for this name only
  
  star_matrix <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = 1,
    dimnames = list(
      rownames(heatmap_avg),
      name
    )
  )

  res <- results_list[[name]]

  common <- intersect(rownames(res), rownames(star_matrix))

  padj <- res[common, "padj"]

  star_matrix[common[!is.na(padj) & padj < 0.001], name] <- "***"
  star_matrix[common[!is.na(padj) & padj < 0.01 & padj >= 0.001], name] <- "**"
  star_matrix[common[!is.na(padj) & padj < 0.05 & padj >= 0.01], name] <- "*"


  # make stars match heatmap dimensions
  stars <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = ncol(heatmap_avg),
    dimnames = dimnames(heatmap_avg)
  )

  stars[, "MW"] <- star_matrix[, name]


  ## save heatmap
  
  pdf(paste0(name, "_hm2.pdf"), width = 5, height = 12)

  pheatmap(
    heatmap_avg,
    display_numbers = stars,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    show_rownames = TRUE,
    fontsize_col = 12,
    fontsize_row = 10,
    cellwidth = 10,
    cellheight = 10,
    main = name
  )

  dev.off()
}
```


```{r}
# heatmap stuffs:
#   color = colorRampPalette(c("lightblue", "yellow", "red"))(100),
```


```{r}
# # 1. Transform the data to stabilize variance
# vst_data <- varianceStabilizingTransformation(dds_sub, blind = FALSE)
# vst_matrix <- assay(vst_data)
# 
# # make list w significant genes
# sig_genes_all <- unique(unlist(
#   lapply(results_list, function(res) {
#     rownames(subset(as.data.frame(res), padj < 0.05))
#   })
# ))

# # 2. Filter for genes that are significant (e.g., padj < 0.05) in WT OR Mutant
# sig_genes_WT <- rownames(subset(res_WT, padj < 0.05))
# sig_genes_Mut <- rownames(subset(res_Mutant, padj < 0.05))
# all_sig_genes <- unique(c(sig_genes_WT, sig_genes_Mut))

# heatmap_matrix <- vst_matrix[all_sig_genes, ]

# 3. Create metadata annotation frame for the top of the heatmap
annotation_df <- data.frame(
  Genotype = your_metadata$Genotype,
  Condition = your_metadata$Condition
)
rownames(annotation_df) <- colnames(vst_matrix)

# 4. Plot Heatmap
pheatmap(heatmap_matrix, 
         annotation_col = annotation_df,
         scale = "row",              # Centers and scales gene expression for visual contrast
         show_rownames = FALSE,      # Turn TRUE if you have few genes and want to read them
         clustering_distance_cols = "euclidean") 


# 2. Select top variable genes (e.g., top 50 genes with highest variance)
vst_matrix <- assay(vst_data)
top_genes <- head(order(rowVars(vst_matrix), decreasing = TRUE), 50)
heatmap_matrix <- vst_matrix[top_genes, ]

# 3. Plot the heatmap (automatically clusters similar samples and genes)
pheatmap(heatmap_matrix, 
         annotation_col = your_metadata["Condition"], # Adds condition color bars
         scale = "row",                               # Normalizes across rows for clear color contrast
         show_rownames = TRUE)

```


## 3. Bar Plots

### 3.1. Analysis differences btwn genotypes, across all segments
```{r}
# set kn as reference:
dat_meta$genotype <- factor(dat_meta$genotype)
dat_meta$genotype <- relevel(dat_meta$genotype, ref = "kn")

# global analysis (all segments toghether)
## run DESeq2:
dds <- DESeqDataSetFromMatrix(
  countData = dat_subset,
  colData = dat_meta,
  design = ~ segment + genotype + condition_group + genotype:condition_group
)

dds <- DESeq(dds)

## create objects for lox-kn and haci-kn comparison
res_lox_vs_kn <- results(
  dds,
  name = "genotypelox.condition_groupw"
)

res_haci_vs_kn <- results(
  dds,
  name = "genotypehaci.condition_groupw"
)

## filter for significant genes:
sig_lox_vs_kn <- res_lox_vs_kn[
  !is.na(res_lox_vs_kn$padj) &
  res_lox_vs_kn$padj < 0.05,
]

sig_haci_vs_kn <- res_haci_vs_kn[
  !is.na(res_haci_vs_kn$padj) &
  res_haci_vs_kn$padj < 0.05,
]
```

### 3.2. Segment-specific analysis
```{r}
# segment-specific analysis (segments separately)
results_genotype <- list()

for (s in unique(dat_meta$segment)) {

  keep <- dat_meta$segment == s

  dds_sub <- DESeqDataSetFromMatrix(
    countData = dat_subset[, keep],
    colData = dat_meta[keep, ],
    design = ~ genotype + condition_group + genotype:condition_group
  )

  dds_sub <- DESeq(dds_sub)

  results_genotype[[paste0(s, "_lox_vs_kn")]] <- results(
    dds_sub,
    name = "genotypelox.condition_groupw"
  )

  results_genotype[[paste0(s, "_haci_vs_kn")]] <- results(
    dds_sub,
    name = "genotypehaci.condition_groupw"
  )
}

# filter for significant genes:
sig_results_genotype <- lapply(
  results_genotype,
  function(x) {
    x[!is.na(x$padj) & x$padj < 0.05, ]
  }
)
```

make gene names:
```{r}
sig_gene_names <- dat_goi$gene_name[
  match(sig_genes, dat_goi$Gene_ID)
]
```

```{r}
# put gene_names into results_list
results_list <- lapply(results_list, function(res) {

  gene_names <- dat_goi$gene_name[
    match(rownames(res), dat_goi$Gene_ID)
  ]

  # Keep original ID if no gene name is found
  gene_names[is.na(gene_names)] <- rownames(res)[is.na(gene_names)]

  rownames(res) <- gene_names

  res
})

# and results_genotype
results_genotype <- lapply(results_genotype, function(res) {

  gene_names <- dat_phytohormone_goi$gene_name[
    match(rownames(res), dat_phytohormone_goi$Gene_ID_v5)
  ]

  # Keep Gene ID if no gene name exists
  gene_names[is.na(gene_names)] <- rownames(res)[is.na(gene_names)]

  rownames(res) <- gene_names

  res
})
```


```{r}


lfc_plot <- bind_rows(
  lapply(names(results_list), function(n) {

    res <- as.data.frame(results_list[[n]])

    res %>%
      rownames_to_column("gene_name") %>%
      filter(gene_name %in% sig_genes) %>%
      mutate(sample = n)

  })
)
```

```{r}
lfc_plot1 <- lfc_plot |>
  separate(sample,
           into = c("genotype", "segment"),
           sep = "_")
```

```{r}
# this plots for each segment only genes that are significant in that segment :)
segments <- c("S1", "S2", "S3")

for (s in segments) {

  ## genes significant in either comparison
  sig_lox <- rownames(
    results_genotype[[paste0(s, "_lox_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_lox_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_lox_vs_kn")]]$padj < 0.05,
    ]
  )

  sig_haci <- rownames(
    results_genotype[[paste0(s, "_haci_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_haci_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_haci_vs_kn")]]$padj < 0.05,
    ]
  )

  sig_genes <- union(sig_lox, sig_haci)

  ## keep only those genes for this segment
  plot_data <- lfc_plot1 %>%
    filter(
      segment == s,
      gene_name %in% sig_genes
    )

  ## genotype order
  plot_data$genotype <- factor(
    plot_data$genotype,
    levels = c("kn", "haci", "lox")
  )

  ## preserve gene order
  plot_data$gene <- factor(
    plot_data$gene_name,
    levels = sig_genes
  )

  p <- ggplot(
    plot_data,
    aes(
      x = gene_name,
      y = log2FoldChange,
      fill = genotype
    )
  ) +
    geom_col(
      position = position_dodge(0.8),
      width = 0.7
    ) +
    geom_errorbar(
      aes(
        ymin = log2FoldChange - lfcSE,
        ymax = log2FoldChange + lfcSE
      ),
      position = position_dodge(0.8),
      width = 0.25
    ) +
    theme_bw() +
    labs(
      title = paste("Segment", s),
      x = "Gene",
      y = "log2FC (w vs c)"
    ) +
    theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1
      )
    )

  ggsave(
    paste0("stress_response_", s, ".pdf"),
    p,
    width = 10,
    height = 6
  )
}
```

```{r}
# for chatgpt:
resultsNames(dds_S1)
```





Try to fix it: (delete after:)

```{r}
for (s in c("S1", "S2", "S3")) {

  sig_lox <- rownames(
    results_genotype[[paste0(s, "_lox_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_lox_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_lox_vs_kn")]]$padj < 0.05,
    ]
  )

  sig_haci <- rownames(
    results_genotype[[paste0(s, "_haci_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_haci_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_haci_vs_kn")]]$padj < 0.05,
    ]
  )

  print(paste(
    s,
    "lox:", length(sig_lox),
    "haci:", length(sig_haci),
    "combined:", length(union(sig_lox, sig_haci))
  ))
}
```

```{r}
s <- "S1"

sig_lox <- rownames(
  results_genotype[["S1_lox_vs_kn"]][
    results_genotype[["S1_lox_vs_kn"]]$padj < 0.05 &
    !is.na(results_genotype[["S1_lox_vs_kn"]]$padj),
  ]
)

sig_haci <- rownames(
  results_genotype[["S1_haci_vs_kn"]][
    results_genotype[["S1_haci_vs_kn"]]$padj < 0.05 &
    !is.na(results_genotype[["S1_haci_vs_kn"]]$padj),
  ]
)

sig_genes <- union(sig_lox, sig_haci)

length(sig_genes)

sum(sig_genes %in% lfc_plot$gene_name)
```
```{r}
lfc_plot1 %>%
  filter(segment == "S1") %>%
  distinct(gene_name) %>%
  nrow()
```
```{r}
lfc_plot1 %>%
  filter(segment == "S1",
         gene_name %in% sig_genes) %>%
  distinct(gene_name) %>%
  nrow()
```
```{r}
length(sig_genes)
```
```{r}
any(duplicated(rownames(results_list[["kn_S1"]])))
sum(duplicated(rownames(results_genotype[["S1_lox_vs_kn"]])))
```
```{r}
length(sig_genes)

sum(sig_genes %in% lfc_plot$gene_name)
```
```{r}
for (s in c("S1", "S2", "S3")) {

  sig_lox <- rownames(
    results_genotype[[paste0(s, "_lox_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_lox_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_lox_vs_kn")]]$padj < 0.05,
    ]
  )

  sig_haci <- rownames(
    results_genotype[[paste0(s, "_haci_vs_kn")]][
      !is.na(results_genotype[[paste0(s, "_haci_vs_kn")]]$padj) &
      results_genotype[[paste0(s, "_haci_vs_kn")]]$padj < 0.05,
    ]
  )

  sig_genes <- union(sig_lox, sig_haci)

  cat(
    s,
    "significant genes:",
    length(sig_genes),
    "\n"
  )

  cat(
    "found in lfc_plot:",
    sum(sig_genes %in% lfc_plot1$gene_name),
    "\n\n"
  )
}
```
```{r}
results_genotype_ID <- results_genotype
results_list_ID <- results_list
```

```{r}
lfc_plot <- bind_rows(
  lapply(names(results_list_ID), function(n) {

    res <- as.data.frame(results_list_ID[[n]])

    res %>%
      tibble::rownames_to_column("Gene_ID") %>%
      mutate(sample = n)

  })
)
```

```{r}
head(lfc_plot$Gene_ID)
```

```{r}
sig_genes <- unique(unlist(
  lapply(
    results_genotype_ID,
    function(x) {
      rownames(
        x[!is.na(x$padj) & x$padj < 0.05, ]
      )
    }
  )
))
```


```{r}
lfc_plot <- lfc_plot %>%
  separate(
    sample,
    into = c("genotype", "segment"),
    sep = "_"
  )
```

```{r}
lfc_plot$gene_name <- dat_phytohormone_goi$gene_name[
  match(lfc_plot$Gene_ID,
        dat_phytohormone_goi$Gene_ID_v5)
]

lfc_plot$gene_name[is.na(lfc_plot$gene_name)] <- lfc_plot$Gene_ID[
  is.na(lfc_plot$gene_name)
]
```

```{r}
for (s in c("S1", "S2", "S3")) {

  sig_genes_s <- unique(unlist(
    lapply(
      results_genotype_ID[grep(s, names(results_genotype_ID))],
      function(x) {
        rownames(x[!is.na(x$padj) & x$padj < 0.05, ])
      }
    )
  ))

  cat(
    s,
    "significant:",
    length(sig_genes_s),
    "found:",
    sum(sig_genes_s %in% lfc_plot$Gene_ID),
    "\n"
  )
}
```

```{r}
dim(lfc_plot)

head(lfc_plot)

table(lfc_plot$sample)
```
```{r}
length(rownames(results_list[["kn_S1"]]))

length(rownames(results_list[["lox_S1"]]))

length(rownames(results_list[["haci_S1"]]))
```

```{r}
sum(is.na(rownames(results_list[["kn_S1"]])))
```

```{r}
sum(is.na(results_list[["kn_S1"]]$log2FoldChange))
```

```{r}
table(is.na(results_list[["kn_S1"]]$log2FoldChange))
```
```{r}
# check if na are in significant genes:
sig_S1 <- union(
  rownames(results_genotype[["S1_lox_vs_kn"]][
    !is.na(results_genotype[["S1_lox_vs_kn"]]$padj) &
    results_genotype[["S1_lox_vs_kn"]]$padj < 0.05,
  ]),
  rownames(results_genotype[["S1_haci_vs_kn"]][
    !is.na(results_genotype[["S1_haci_vs_kn"]]$padj) &
    results_genotype[["S1_haci_vs_kn"]]$padj < 0.05,
  ])
)

lfc_S1 <- results_list[["kn_S1"]]

sig_S1[is.na(lfc_S1[sig_S1, "log2FoldChange"])]
```

























```{r}
# plot segments:
# this plots it for all genes that are significant in any of the segments for all segments (which is not what I want)
for (s in unique(lfc_plot1$segment)) {
  plot_data <- lfc_plot1 %>%
    filter(segment == s)
  
  plot_data$genotype <- factor(
    plot_data$genotype,
    levels = c("kn", "haci", "lox")
  )

  p <-ggplot(
    plot_data,
      aes(
        x = gene_name,
        y = log2FoldChange,
        fill = genotype
      )
    ) +
    geom_col(
      position = position_dodge(0.8),
      width = 0.7
    ) +
    geom_errorbar(
      aes(
        ymin = log2FoldChange - lfcSE,
        ymax = log2FoldChange + lfcSE
      ),
      position = position_dodge(0.8),
      width = 0.25
    ) +
    theme_bw() +
    labs(
      title = paste("Segment", s),
      y = "log2FC (w vs c)",
      x = "Gene"
    ) +
    theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1
      )
    )

  print(p)
  ggsave(
    filename = paste0("stress_response_", s, ".pdf"),
    plot = p,
    width = 10,
    height = 6
  )
}
```





Make a df out of this:
```{r}

sig_results_genotype <- lapply(
  results_genotype,
  function(x) {
    as.data.frame(x) %>%
      filter(!is.na(padj) & padj < 0.05)
  }
)

plot_data <- bind_rows(
  lapply(names(sig_results_genotype), function(n) {
    sig_results_genotype[[n]] %>%
      rownames_to_column("gene") %>%
      mutate(comparison = n)
  })
)
```

add gene names:
```{r}
plot_data$gene_name <- dat_phytohormone_goi$gene_name[
  match(plot_data$gene, dat_phytohormone_goi$Gene_ID_v5)
]

plot_data$gene_name[is.na(plot_data$gene_name)] <- plot_data$gene[
  is.na(plot_data$gene_name)
]
```

Get significant genes, to go back to result list and make barplot:
```{r}
sig_genes <- unique(unlist(
  lapply(sig_results_genotype, rownames)
))

# create plotting df:

lfc_plot <- bind_rows(
  lapply(names(results_list), function(n) {

    res <- as.data.frame(results_list[[n]])

    res %>%
      rownames_to_column("gene") %>%
      filter(gene %in% sig_genes) %>%
      mutate(sample = n)

  })
)

length(sig_genes)

head(sig_genes)

head(rownames(results_list[["kn_S1"]]))

sum(sig_genes %in% rownames(results_list[["kn_S1"]]))
```





Plot the delta lfc: (I don't need this actly)
```{r}
ggplot(
  plot_data,
  aes(
    x = gene_name,
    y = log2FoldChange
  )
) +
  geom_col() +
  facet_wrap(~ comparison, scales = "free_x") +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  ) +
  ylab("Difference in stress response\n(log2FC genotype interaction)") +
  xlab("")
```
















Comparing log-fold-change between genotypes: Is there a difference between genotypes in the effect wounding has (compared to c)

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = dat_subset_ph,
  colData = dat_meta,
  design = ~ genotype + segment + condition_group + genotype:condition_group
)

dds <- DESeq(dds)

res_interaction <- results(
  dds,
  name = "genotypekn.condition_groupw"
)

summary(res_interaction)

head(res_interaction[order(res_interaction$padj), ])

sum(!is.na(res_interaction$padj))

sig <- res_interaction[
  !is.na(res_interaction$padj) &
  res_interaction$padj < 0.05 &
  abs(res_interaction$log2FoldChange) > 1,
]

sig$gene_name <- dat_phytohormone_goi$gene_name[
  match(rownames(sig), dat_phytohormone_goi$Gene_ID_v5)
]


```


```{r}
sig_genes <- rownames(sig)
```


















```{r}

lfc_df <- do.call(rbind,
                  lapply(names(results_list), function(name) {
                    
                    res <- as.data.frame(results_list[[name]])
                    
                    data.frame(
                      gene_name = rownames(res),
                      comparison = name,
                      log2FC = res$log2FoldChange,
                      se = res$lfcSE,
                      pvalue = res$pvalue,
                      padj = res$padj
                    )
                  }))
```

```{r}
lfc_df1 <- lfc_df |>
  left_join(dat_phytohormone_goi, by = "gene_name")

lfc_df2 <- lfc_df1 %>%
  separate(comparison,
           into = c("genotype", "segment"),
           sep = "_")

# subset by segment:

lfc_s1 <- subset(lfc_df2, segment == "S1")
lfc_s2 <- subset(lfc_df2, segment == "S2")
lfc_s3 <- subset(lfc_df2, segment == "S3")

# 

```

```{r}
unique((lfc_df2$category))
```


```{r}

gene_plot <- "ABA-related genes"

plot_df <- lfc_df2 %>%
  filter(category == gene_plot)

# plot_df <- subset(lfc_df2, gene_name %in% c("TPS3", "CYP92C5"))



ggplot(plot_df, aes(x = gene_name, y = log2FC, fill = genotype)) +
  geom_bar(position = "dodge", stat = "identity", color = "black") +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin=log2FC-se*1.96, ymax= log2FC + se*1.96), 
                width=0,
                position = position_dodge(width = 0.9)) +
  facet_wrap(~ segment) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 1)
) +
  ylab("log2 fold change (Treat vs Ctrl)") +
  xlab("")
```

```{r}

gene_plot <- "JA"

plot_df <- lfc_df2 %>%
  filter(category == gene_plot)

ggplot(plot_df, aes(x = gene_name, y = log2FC, fill = genotype)) +
  geom_bar(position = "dodge", stat = "identity", color = "black") +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin=log2FC-se*1.96, ymax= log2FC + se*1.96), 
                width=0,
                position = position_dodge(width = 0.9)) +
  facet_wrap(~ segment) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 1)
) +
  ylab("log2 fold change (Treat vs Ctrl)") +
  xlab("")
```


```{r}

gene_plot <- "Transcription factors / JA signaling"

plot_df <- lfc_df2 %>%
  filter(category == gene_plot)

ggplot(plot_df, aes(x = gene_name, y = log2FC, fill = genotype)) +
  geom_bar(position = "dodge", stat = "identity", color = "black") +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin=log2FC-se*1.96, ymax= log2FC + se*1.96), 
                width=0,
                position = position_dodge(width = 0.9)) +
  facet_wrap(~ segment) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 1)
) +
  ylab("log2 fold change (Treat vs Ctrl)") +
  xlab("")
```


```{r}

gene_plot <- "Benzoxazinoids"

plot_df <- lfc_df2 %>%
  filter(category == gene_plot)

ggplot(plot_df, aes(x = gene_name, y = log2FC, fill = genotype)) +
  geom_bar(position = "dodge", stat = "identity", color = "black") +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin=log2FC-se*1.96, ymax= log2FC + se*1.96), 
                width=0,
                position = position_dodge(width = 0.9)) +
  facet_wrap(~ segment) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 1)
) +
  ylab("log2 fold change (Treat vs Ctrl)") +
  xlab("")
```


# ChatGPT stuff not looped:


------
ChatGPT stuff:
```{r}
colnames(dat_meta)
head(dat_meta)
head(colnames(dat_raw_counts))
rownames(dat_meta)
```

```{r}
unique(dat_meta$genotype)
unique(dat_meta$condition_code)
unique(dat_meta$condition_group)
unique(dat_meta$segment)

table(
  dat_meta$genotype,
  dat_meta$condition_group,
  dat_meta$segment
)
```



```{r}
# make rownames dat_meta match dat_raw_counts
rownames(dat_meta) <- dat_meta$sample_description

# verify:
all(rownames(dat_meta) %in% colnames(dat_raw_counts))
all(rownames(dat_meta) == colnames(dat_raw_counts))

```

```{r}
# check for S1:

# Make sample IDs the rownames
rownames(dat_meta) <- dat_meta$sample_description

# Ensure same order
dat_raw_counts <- dat_raw_counts[, rownames(dat_meta)]

# Factors
dat_meta$genotype <- factor(
  dat_meta$genotype,
  levels = c("kn", "lox", "haci")
)

dat_meta$condition_group <- factor(
  dat_meta$condition_group,
  levels = c("w", "c")
)

dat_meta$segment <- factor(
  dat_meta$segment,
  levels = c("S1", "S2", "S3")
)

# -----------------------
# S1
# -----------------------

meta_S1 <- dat_meta %>%
  filter(segment == "S1")

counts_S1 <- dat_raw_counts[, rownames(meta_S1)] # subsets colnames raw counts that are S1

dds_S1 <- DESeqDataSetFromMatrix(
  countData = round(as.matrix(counts_S1)),
  colData = meta_S1,
  design = ~ genotype + condition_group + genotype:condition_group
)

# Remove very low-count genes
keep <- rowSums(counts(dds_S1) >= 10) >= 2
dds_S1 <- dds_S1[keep, ]

dds_S1 <- DESeq(dds_S1)

resultsNames(dds_S1)
```

```{r}
# subset
res_lox_interaction <- results(
  dds_S1,
  name = "genotypelox.condition_groupc"
)

res_haci_interaction <- results(
  dds_S1,
  name = "genotypehaci.condition_groupc"
)
```

```{r}
# select genes with FDR < 0.05
sig_lox <- rownames(res_lox_interaction)[
  !is.na(res_lox_interaction$padj) &
  res_lox_interaction$padj < 0.05
]

sig_haci <- rownames(res_haci_interaction)[
  !is.na(res_haci_interaction$padj) &
  res_haci_interaction$padj < 0.05
]
```

```{r}
# combine
# sig_genes <- union(sig_lox, sig_haci)  # this keeps all genes detected

sig_genes <- intersect(  # this only keeps genes that are significant and are also in the genes of interest list 1
  union(sig_lox, sig_haci),
  dat_goi$Gene_ID
)


# sig_genes <- intersect(  # this only keeps genes that are significant and are also in the genes of interest list 2 (phytohormones)
#   union(sig_lox, sig_haci),
#   dat_phytohormone_goi$Gene_ID_v5
# )
```

log-fold change:
```{r}
res_WT <- results(
  dds_S1,
  contrast = c(
    "condition_group",
    "c",
    "w"
  )
)

res_lox <- results(
  dds_S1,
  contrast = list(
    c(
      "condition_group_c_vs_w",
      "genotypelox.condition_groupc"
    )
  )
)

res_haci <- results(
  dds_S1,
  contrast = list(
    c(
      "condition_group_c_vs_w",
      "genotypehaci.condition_groupc"
    )
  )
)
```

```{r}
# plotting data frame:
plot_S1 <- data.frame(
  gene = sig_genes,
  WT = res_WT[sig_genes, "log2FoldChange"],
  lox = res_lox[sig_genes, "log2FoldChange"],
  haci = res_haci[sig_genes, "log2FoldChange"]
)

plot_S1_1 <- plot_S1 |>
  mutate(Gene_ID = gene) |>
  left_join(dat_goi, gene_name, by = "Gene_ID")
```

```{r}
# long df
plot_S1_long <- plot_S1_1 %>%
  pivot_longer(
    cols = c(WT, lox, haci),
    names_to = "genotype",
    values_to = "log2FC"
  )
```

```{r}
# plot:
ggplot(
  plot_S1_long,
  aes(
    x = gene_name,
    y = log2FC,
    fill = genotype
  )
) +
  geom_col(
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.5
  ) +
  theme_classic() +
  labs(
    title = "S1",
    x = NULL,
    y = "log2 fold change (C vs W)",
    fill = "Genotype"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5
    )
  )
```

# All genes

```{r}
genotypes <- unique(dat_meta$genotype)
segments <- unique(dat_meta$segment)

results_list <- list()
dds_list <- list()

for(g in genotypes){
  for(s in segments){

    keep <- dat_meta$genotype == g &
            dat_meta$segment == s

    dds_sub <- DESeqDataSetFromMatrix(
      countData = dat_raw_counts[, keep],
      colData = dat_meta[keep, ],
      design = ~ condition_group
    )

    dds_sub <- DESeq(dds_sub)

    res <- results(dds_sub,
                   contrast = c("condition_group", "w", "c"))

    results_list[[paste(g, s, sep = "_")]] <- res
    dds_list[[paste(g, s, sep = "_")]] <- dds_sub
  }
}

```








