---
title: "RNAseq local response"
author: "Sara Hoefer"
date: "2026-07-27"
output: html_document
---

This script analyzes RNAseq (transcriptomics) data. It's my first time doing this lol

----------------------------
HOW TO USE THIS SCRIPT
Always run Sections 1. Packages, 2. Functions, 3. Data
Sections 4. Differential Expression and 5. Functional Analyses should run independently of each other.

---------------------------

To write section 4 of this script, I heavily depended on AI. Section 5.1. is based off this v helpful github post:
https://aseetharam.github.io/bioinformatics/run-go-enrichment-maize/

The B73_GO.out data was downloaded from (you need to use the command line):
https://download.maizegdb.org/GeneFunction_and_Expression/Pannzer_GO_Terms/B73_GO.out


# 1. Packages
```{r}
library(pheatmap)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(RColorBrewer) # to change colors in pheatmap
library(tidyverse)
library(tibble)
library(scales)

## to install DESeq2:
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("DESeq2")

# bioconductor:

# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("topGO")

library(topGO)
```

# 2. Functions
```{r}
se <- function (x) {sd (x,na.rm =TRUE) / sqrt (length(na.exclude (x)))}
```


# 3. Data
## 3.1. Load data

```{r}
dat_raw_counts <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/01_data/within_leaf_analysis_ready_gene_counts.tsv",
                             header = T, row.names = 1, sep = "\t")

dat_fpkm <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/01_data/within_leaf_analysis_ready_gene_fpkm_all_samples.tsv",
                             header = T, row.names = 1, sep = "\t")

dat_goi <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/01_data/tristan_heatmap_gene_categories.csv",
                      header = T, sep = ",")  # first list of genes involved in defense from Tristan

dat_phytohormone_goi <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/01_data/phytohormone_candidates_gene_list.csv",
                                   header = T, sep = ",")  # second list of genes involved in defense from Tristan

dat_meta <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/01_data/within_leaf_analysis_ready_info.csv",
                             header = T, sep = ",")  # metadata

dat_b73_go <- read.table("C:/Users/sarah/OneDrive/Dokumente/01_Uni/00_PhD/01_Thesis/03_local_response_RNAseq/B73_GO.out",
                         quote = '"',
                             header = T, sep = "\t",
                         colClasses = c('goid' = 'character', 'qpid' = 'character'))
```

## 3.2. Wrangle data

```{r}
# create col "segment" in metadata

dat_meta$segment <- sub(".*(S.).*", "\\1", dat_meta$sample_description)


# change phytohormone Gene_ID_v5 to Gene_ID

dat_ph_goi <- dat_phytohormone_goi |>
  mutate(Gene_ID = Gene_ID_v5)
```

# 4. Differential Expression Analyses:
## 4.1. Heatmap
### 4.1.1 DESeq2 for heatmap

```{r}

# 1. Create DESeq2 object (Replace with your actual data frames)
# countData: matrix of raw gene counts
# colData: dataframe with sample metadata, last condition is the default comparison (so it's treat vs. ctrl)

genotypes <- unique(dat_meta$genotype)
segments <- unique(dat_meta$segment)

results_list <- list()
dds_list <- list()

for(g in genotypes){
  for(s in segments){

    keep <- dat_meta$genotype == g &
            dat_meta$segment == s

    dds_sub <- DESeqDataSetFromMatrix(
      countData = dat_subset_ph[, keep],
      colData = dat_meta[keep, ],
      design = ~ condition_group
    )

    dds_sub <- DESeq(dds_sub)

    res <- results(dds_sub,
                   contrast = c("condition_group", "w", "c"))

    results_list[[paste(g, s, sep = "_")]] <- res
    dds_list[[paste(g, s, sep = "_")]] <- dds_sub
  }
}

```

```{r}
# create matrix for heatmap

for (name in c(
  "kn_S1", "kn_S2", "kn_S3",
  "lox_S1", "lox_S2", "lox_S3",
  "haci_S1", "haci_S2", "haci_S3"
)) {

  # DESeq2 results
  res <- as.data.frame(results_list[[name]])

  # VST transform
  vst_data <- varianceStabilizingTransformation(
    dds_list[[name]],
    blind = FALSE
  )

  vst_matrix <- assay(vst_data)

  # scale each gene
  heatmap_matrix <- t(scale(t(vst_matrix)))

  # get treatment labels
  condition <- colData(dds_list[[name]])$condition_group

  # average replicates
  heatmap_avg <- sapply(
    levels(condition),
    function(x) {
      rowMeans(heatmap_matrix[, condition == x, drop = FALSE])
    }
  )

  # reorder rows according to dat_phytohormone_goi order
  row_order <- match(dat_phytohormone_goi$Gene_ID_v5, rownames(heatmap_avg))
  row_order <- row_order[!is.na(row_order)]

  heatmap_avg <- heatmap_avg[row_order, ]

  # replace row names with gene names
  gene_names <- dat_phytohormone_goi$gene_name[
    match(rownames(heatmap_avg), dat_phytohormone_goi$Gene_ID_v5)
  ]

  rownames(heatmap_avg) <- gene_names
  colnames(heatmap_avg) <- c("Control", "MW")


  ## star matrix for this name only
  
  star_matrix <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = 1,
    dimnames = list(
      rownames(heatmap_avg),
      name
    )
  )

  res <- results_list[[name]]

  common <- intersect(rownames(res), rownames(star_matrix))

  padj <- res[common, "padj"]

  star_matrix[common[!is.na(padj) & padj < 0.001], name] <- "***"
  star_matrix[common[!is.na(padj) & padj < 0.01 & padj >= 0.001], name] <- "**"
  star_matrix[common[!is.na(padj) & padj < 0.05 & padj >= 0.01], name] <- "*"


  # make stars match heatmap dimensions
  stars <- matrix(
    "",
    nrow = nrow(heatmap_avg),
    ncol = ncol(heatmap_avg),
    dimnames = dimnames(heatmap_avg)
  )

  stars[, "MW"] <- star_matrix[, name]


  ## save heatmap
  
  # pdf(paste0(name, "_hm2.pdf"), width = 5, height = 12)

  pheatmap(
    heatmap_avg,
    display_numbers = stars,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    show_rownames = TRUE,
    fontsize_col = 12,
    fontsize_row = 10,
    cellwidth = 10,
    cellheight = 10,
    main = name
  )

  # dev.off()
}
```

## 4.2. Barplots log-fold change

### 4.2.1. Subset of genes

Loop for everything from ChatGPT:

```{r}

# ---------------------------------------------------------
# Prepare metadata
# ---------------------------------------------------------

rownames(dat_meta) <- dat_meta$sample_description

# Make sure count matrix follows metadata order
dat_raw_counts <- dat_raw_counts[, rownames(dat_meta)]

# Factors
dat_meta$genotype <- factor(
  dat_meta$genotype,
  levels = c("kn", "lox", "haci")
)

dat_meta$condition_group <- factor(
  dat_meta$condition_group,
  levels = c("w", "c")
)

dat_meta$segment <- factor(
  dat_meta$segment,
  levels = c("S1", "S2", "S3")
)


# ---------------------------------------------------------
# Function
# ---------------------------------------------------------

make_segment_plot <- function(segment_name,
                              counts,
                              metadata,
                              goi,
                              padj_cutoff = 0.05) {
  
  # -------------------------------------------------------
  # 1. Subset metadata and counts for this segment
  # -------------------------------------------------------
  
  meta_seg <- metadata %>%
    filter(segment == segment_name)
  
  counts_seg <- counts[, rownames(meta_seg)]
  
  
  # -------------------------------------------------------
  # 2. Create DESeq2 object
  # -------------------------------------------------------
  
  dds <- DESeqDataSetFromMatrix(
    countData = round(as.matrix(counts_seg)),
    colData = meta_seg,
    design = ~ genotype + condition_group +
              genotype:condition_group
  )
  
  
  # -------------------------------------------------------
  # 3. Filter low-count genes
  # -------------------------------------------------------
  
  keep <- rowSums(counts(dds) >= 10) >= 2
  dds <- dds[keep, ]
  
  
  # -------------------------------------------------------
  # 4. Run DESeq2
  # -------------------------------------------------------
  
  dds <- DESeq(dds)
  
  
  # -------------------------------------------------------
  # 5. Interaction tests
  #
  # Does the C-vs-W response differ between:
  #   WT vs lox
  #   WT vs haci
  # -------------------------------------------------------
  
  res_lox_interaction <- results(
    dds,
    name = "genotypelox.condition_groupc"
  )
  
  res_haci_interaction <- results(
    dds,
    name = "genotypehaci.condition_groupc"
  )
  
  
  # -------------------------------------------------------
  # 6. Significant interaction genes
  # -------------------------------------------------------
  
  sig_lox <- rownames(res_lox_interaction)[
    !is.na(res_lox_interaction$padj) &
      res_lox_interaction$padj < padj_cutoff
  ]
  
  sig_haci <- rownames(res_haci_interaction)[
    !is.na(res_haci_interaction$padj) &
      res_haci_interaction$padj < padj_cutoff
  ]
  
  
  # Significant in WT-vs-lox OR WT-vs-haci
  sig_genes <- union(
    sig_lox,
    sig_haci
  )
  
  
  # -------------------------------------------------------
  # 7. Restrict to genes of interest
  # -------------------------------------------------------
  
  sig_genes <- intersect(
    sig_genes,
    goi$Gene_ID
  )
  
  
  # -------------------------------------------------------
  # 8. Get C vs W effect for WT
  # -------------------------------------------------------
  
  res_WT <- results(
    dds,
    contrast = c(
      "condition_group",
      "c",
      "w"
    )
  )
  
  
  # -------------------------------------------------------
  # 9. Get C vs W effect for lox
  # -------------------------------------------------------
  
  res_lox <- results(
    dds,
    contrast = list(
      c(
        "condition_group_c_vs_w",
        "genotypelox.condition_groupc"
      )
    )
  )
  
  
  # -------------------------------------------------------
  # 10. Get C vs W effect for haci
  # -------------------------------------------------------
  
  res_haci <- results(
    dds,
    contrast = list(
      c(
        "condition_group_c_vs_w",
        "genotypehaci.condition_groupc"
      )
    )
  )
  
  
  # -------------------------------------------------------
  # 11. Create plotting dataframe
  #
  # Keep both log2FC and lfcSE
  # -------------------------------------------------------
  
  plot_data <- bind_rows(
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "WT",
      log2FC = res_WT[sig_genes, "log2FoldChange"],
      lfcSE = res_WT[sig_genes, "lfcSE"]
    ),
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "lox",
      log2FC = res_lox[sig_genes, "log2FoldChange"],
      lfcSE = res_lox[sig_genes, "lfcSE"]
    ),
    
    data.frame(
      Gene_ID = sig_genes,
      genotype = "haci",
      log2FC = res_haci[sig_genes, "log2FoldChange"],
      lfcSE = res_haci[sig_genes, "lfcSE"]
    )
  )
  
  # ------------------------------------------------------- 
  # 12. Add interaction p-values to plotting data
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    padj_lox_vs_WT = res_lox_interaction[Gene_ID, "padj"],
    padj_haci_vs_WT = res_haci_interaction[Gene_ID, "padj"]
  )
  
  
  # -------------------------------------------------------
  # 13. Create a significance label:
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    significance = case_when(
      genotype == "lox" & padj_lox_vs_WT < 0.001  ~ "***",
      genotype == "lox" & padj_lox_vs_WT < 0.01   ~ "**",
      genotype == "lox" & padj_lox_vs_WT < 0.05   ~ "*",
      
      genotype == "haci" & padj_haci_vs_WT < 0.001  ~ "***",
      genotype == "haci" & padj_haci_vs_WT < 0.01   ~ "**",
      genotype == "haci" & padj_haci_vs_WT < 0.05   ~ "*",
      
      TRUE ~ ""
    )
  )
  
  # -------------------------------------------------------
  # 14. Calculate vjust in data:
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
  mutate(
    star_y = ifelse(
      log2FC >= 0,
      log2FC + 1.96 * lfcSE,
      log2FC - 1.96 * lfcSE
    ),
    star_vjust = ifelse(
      log2FC >= 0,
      -0.3,
      1.3
    )
  )
  
  
  # -------------------------------------------------------
  # 15. Add gene names
  # -------------------------------------------------------
  
  plot_data <- plot_data %>%
    left_join(
      goi %>%
        select(Gene_ID, gene_name),
      by = "Gene_ID"
    )
  
  
  # -------------------------------------------------------
  # 16. Plot
  # -------------------------------------------------------
  
  p <- ggplot(
    plot_data,
    aes(
      x = gene_name,
      y = log2FC,
      fill = genotype
    )
  ) +
    
    geom_col(
      position = position_dodge(width = 0.8)
    ) +
    
    geom_errorbar(
      aes(
        ymin = log2FC - 1.96 * lfcSE,
        ymax = log2FC + 1.96 * lfcSE
      ),
      position = position_dodge(width = 0.8),
      width = 0.25
    ) +
    
    geom_text(
      aes(
        label = significance,
        y = star_y,
        vjust = star_vjust
      ),
      position = position_dodge(width = 0.8),
      size = 4
    ) +
    
    geom_hline(
      yintercept = 0,
      linewidth = 0.5
    ) +
    
    theme_classic() +
    
    labs(
      title = segment_name,
      x = NULL,
      y = "log2 fold change (C vs W)",
      fill = "Genotype"
    ) +
    
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5
      )
    )
  
  
  # -------------------------------------------------------
  # 17. Return everything
  # -------------------------------------------------------
  
  return(
    list(
      dds = dds,
      significant_genes = sig_genes,
      interaction_lox = res_lox_interaction,
      interaction_haci = res_haci_interaction,
      plot_data = plot_data,
      plot = p
    )
  )
}
```


```{r}
# directly change input here

S1 <- make_segment_plot(
  segment_name = "S1",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)

S2 <- make_segment_plot(
  segment_name = "S2",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)

S3 <- make_segment_plot(
  segment_name = "S3",
  counts = dat_raw_counts,
  metadata = dat_meta,
  goi = dat_ph_goi
)
```

```{r}
fig_S1 <- S1$plot
fig_S2 <- S2$plot
fig_S3 <- S3$plot

fig_S1
fig_S2
fig_S3
```

```{r}
# save plots:

# pdf(
#   "S1_barplot2.pdf",
#   width = 12,
#   height = 5
# )
# 
# print(fig_S1)
# 
# dev.off()
```


# 5. Functional interpretation

Make lists of genes for GO-Term analyses

```{r}
# library(dplyr)
# library(tibble)


# ---------------------------------------------------------
# Prepare metadata
# ---------------------------------------------------------

rownames(dat_meta) <- dat_meta$sample_description

dat_raw_counts <- dat_raw_counts[, rownames(dat_meta), drop = FALSE]

dat_meta$genotype <- factor(
  dat_meta$genotype,
  levels = c("kn", "lox", "haci")
)

dat_meta$condition_group <- factor(
  dat_meta$condition_group,
  levels = c("w", "c")
)

dat_meta$segment <- factor(
  dat_meta$segment,
  levels = c("S1", "S2", "S3")
)


# ---------------------------------------------------------
# Function
# ---------------------------------------------------------

run_segment_interaction <- function(
    segment_name,
    counts,
    metadata,
    padj_cutoff = 0.05
) {
  
  # Subset metadata
  meta_seg <- metadata[
    as.character(metadata$segment) == segment_name,
    ,
    drop = FALSE
  ]
  
  # Keep only samples present in counts
  meta_seg <- meta_seg[
    rownames(meta_seg) %in% colnames(counts),
    ,
    drop = FALSE
  ]
  
  # Subset counts
  counts_seg <- counts[, rownames(meta_seg), drop = FALSE]
  
  # Ensure identical sample order
  counts_seg <- counts_seg[, rownames(meta_seg), drop = FALSE]
  
  # Drop unused factor levels
  meta_seg$genotype <- droplevels(meta_seg$genotype)
  meta_seg$condition_group <- droplevels(meta_seg$condition_group)
  
  # -------------------------------------------------------
  # DESeq2 object
  # -------------------------------------------------------
  
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(as.matrix(counts_seg)),
    colData = meta_seg,
    design = ~ genotype + condition_group +
      genotype:condition_group
  )
  
  # -------------------------------------------------------
  # Filter low-count genes
  # -------------------------------------------------------
  
  keep <- rowSums(DESeq2::counts(dds) >= 10) >= 2
  
  dds <- dds[keep, ]
  
  # -------------------------------------------------------
  # Run DESeq2
  # -------------------------------------------------------
  
  dds <- DESeq2::DESeq(dds)
  
  # -------------------------------------------------------
  # lox10 vs WT interaction
  # -------------------------------------------------------
  
  res_lox <- DESeq2::results(
    dds,
    name = "genotypelox.condition_groupc"
  )
  
  # -------------------------------------------------------
  # haci vs WT interaction
  # -------------------------------------------------------
  
  res_haci <- DESeq2::results(
    dds,
    name = "genotypehaci.condition_groupc"
  )
  
  # -------------------------------------------------------
  # Convert to data frames
  # -------------------------------------------------------
  
  res_lox_df <- as.data.frame(res_lox)
  res_lox_df$Gene_ID <- rownames(res_lox_df)
  rownames(res_lox_df) <- NULL
  
  res_lox_df$segment <- segment_name
  res_lox_df$comparison <- "lox10_vs_WT"
  
  res_haci_df <- as.data.frame(res_haci)
  res_haci_df$Gene_ID <- rownames(res_haci_df)
  rownames(res_haci_df) <- NULL
  
  res_haci_df$segment <- segment_name
  res_haci_df$comparison <- "haci_vs_WT"
  
  # -------------------------------------------------------
  # Significant genes
  # -------------------------------------------------------
  
  sig_lox <- res_lox_df[
    !is.na(res_lox_df$padj) &
      res_lox_df$padj < padj_cutoff,
    ,
    drop = FALSE
  ]
  
  sig_haci <- res_haci_df[
    !is.na(res_haci_df$padj) &
      res_haci_df$padj < padj_cutoff,
    ,
    drop = FALSE
  ]
  
  # -------------------------------------------------------
  # Positive interaction
  # -------------------------------------------------------
  
  lox10_positive <- sig_lox$Gene_ID[
    sig_lox$log2FoldChange > 0
  ]
  
  haci_positive <- sig_haci$Gene_ID[
    sig_haci$log2FoldChange > 0
  ]
  
  # -------------------------------------------------------
  # Negative interaction
  # -------------------------------------------------------
  
  lox10_negative <- sig_lox$Gene_ID[
    sig_lox$log2FoldChange < 0
  ]
  
  haci_negative <- sig_haci$Gene_ID[
    sig_haci$log2FoldChange < 0
  ]
  
  # -------------------------------------------------------
  # Either mutant
  # -------------------------------------------------------
  
  either_mutant <- union(
    sig_lox$Gene_ID,
    sig_haci$Gene_ID
  )
  
  either_mutant_positive <- union(
    lox10_positive,
    haci_positive
  )
  
  either_mutant_negative <- union(
    lox10_negative,
    haci_negative
  )
  
  # -------------------------------------------------------
  # Return
  # -------------------------------------------------------
  
  list(
    dds = dds,
    
    lox10_vs_WT = res_lox_df,
    haci_vs_WT = res_haci_df,
    
    significant_lox10_vs_WT = sig_lox,
    significant_haci_vs_WT = sig_haci,
    
    lox10_positive = lox10_positive,
    lox10_negative = lox10_negative,
    
    haci_positive = haci_positive,
    haci_negative = haci_negative,
    
    either_mutant = either_mutant,
    either_mutant_positive = either_mutant_positive,
    either_mutant_negative = either_mutant_negative
  )
}


# ---------------------------------------------------------
# Run all segments
# ---------------------------------------------------------

segments <- unique(as.character(dat_meta$segment))

DE_results <- lapply(
  segments,
  function(seg) {
    
    message("Running segment: ", seg)
    
    run_segment_interaction(
      segment_name = seg,
      counts = dat_raw_counts,
      metadata = dat_meta,
      padj_cutoff = 0.05
    )
  }
)

names(DE_results) <- segments


# ---------------------------------------------------------
# Gene vectors for topGO
# ---------------------------------------------------------

topGO_genes <- lapply(
  DE_results,
  function(x) {
    
    list(
      lox10_positive = x$lox10_positive,
      lox10_negative = x$lox10_negative,
      
      haci_positive = x$haci_positive,
      haci_negative = x$haci_negative,
      
      either_mutant = x$either_mutant,
      either_mutant_positive = x$either_mutant_positive,
      either_mutant_negative = x$either_mutant_negative
    )
  }
)


# ---------------------------------------------------------
# Individual vectors
# ---------------------------------------------------------

S1_lox10_positive <- topGO_genes$S1$lox10_positive
S1_lox10_negative <- topGO_genes$S1$lox10_negative
S1_haci_positive <- topGO_genes$S1$haci_positive
S1_haci_negative <- topGO_genes$S1$haci_negative

S2_lox10_positive <- topGO_genes$S2$lox10_positive
S2_lox10_negative <- topGO_genes$S2$lox10_negative
S2_haci_positive <- topGO_genes$S2$haci_positive
S2_haci_negative <- topGO_genes$S2$haci_negative

S3_lox10_positive <- topGO_genes$S3$lox10_positive
S3_lox10_negative <- topGO_genes$S3$lox10_negative
S3_haci_positive <- topGO_genes$S3$haci_positive
S3_haci_negative <- topGO_genes$S3$haci_negative


# ---------------------------------------------------------
# Check number of genes
# ---------------------------------------------------------

sapply(
  topGO_genes,
  function(x) sapply(x, length)
)
```




## 5.1. GO-Term analysis

```{r}
# format go-term df:
dat_b73_go1 <- dat_b73_go[(dat_b73_go$ARGOT_PPV > 0.5), ]  # filter to only include genes that have confident ARGOT_score assigned to them

dat_b73_go1$goid <- paste0('GO:', dat_b73_go1$goid)  # Add "GO:" infront of all GO-IDs

dat_b73_go2 <- dat_b73_go1 |>
  separate(qpid, into = c("qpid", "P_isoform"), sep = "_")

panzer_to_golist <- function(panzer_df){
  go_df <- aggregate( goid ~ qpid, data=panzer_df, FUN=c)
  structure(go_df$goid, .Names=go_df$qpid)
}
all_golist <- panzer_to_golist(dat_b73_go2)
```

TopGO needs the full list of genes with a column "myInput" that contains "yes" or "no", yes denotes the genes I would like to run the enrichment analysis on.

1. make a df that is just a list of the genes that are differentially expressed:
```{r}
S1 <- topGO_genes$S1$haci_positive
S2 <- topGO_genes$S2$haci_positive
S3 <- topGO_genes$S3$haci_positive

dat_s1 <- as.data.frame(S1)
dat_s2 <- as.data.frame(S2)
dat_s3 <- as.data.frame(S3)

colnames(dat_s1) <- "genenames"
colnames(dat_s2) <- "genenames"
colnames(dat_s3) <- "genenames"

dat_s1 <- dat_s1 |>
  mutate(myInput = "yes")

dat_s2 <- dat_s2 |>
  mutate(myInput = "yes")

dat_s3 <- dat_s3 |>
  mutate(myInput = "yes")

```


```{r}
# create full list of genes
# gene_names <- names(all_golist)
# myGenes <- as.data.frame(gene_names)
# colnames(myGenes) <- "genenames"
# myGenes <- left_join(myGenes, dat_s2)
# myGenes$myInput <- myGenes$myInput %>%
#   replace_na('no')
# myGenesF <- factor(as.integer(myGenes$myInput == 'yes'))
# names(myGenesF) <- gene_names

# Function that does it for all:
make_topGO_gene_vector <- function(selected_genes, background_genes) {

  # Remove genes that aren't in the GO annotation background
  selected_genes <- intersect(
    selected_genes,
    background_genes
  )

  # Create yes/no vector
  myGenesF <- factor(
    as.integer(background_genes %in% selected_genes)
  )

  names(myGenesF) <- background_genes

  return(myGenesF)
}

myGenesF <- make_topGO_gene_vector(
  topGO_genes$S2$haci_positive,
  gene_names
)
```

Run enrichment analysis:
```{r}
# run_topGO <- function(gene_list, ontology, gene2GO_list){
#   topGO_data <- new("topGOdata", ontology = ontology, allGenes = gene_list,
#                     annot = annFUN.gene2GO, gene2GO = gene2GO_list)
#   fishers_result <- runTest(topGO_data, algorithm = "elim", statistic = "fisher")
#   fishers_table <- GenTable(topGO_data, Fishers = fishers_result, useLevels = TRUE)
#   fishers_table$Ontology <- ontology
#   fishers_table
# }

# Add FDR

run_topGO <- function(gene_list, ontology, gene2GO_list) {

  topGO_data <- new(
    "topGOdata",
    ontology = ontology,
    allGenes = gene_list,
    annot = annFUN.gene2GO,
    gene2GO = gene2GO_list
  )

  fishers_result <- runTest(
    topGO_data,
    algorithm = "elim",
    statistic = "fisher"
  )

  fishers_table <- GenTable(
    topGO_data,
    Fishers = fishers_result,
    useLevels = TRUE
  )

  fishers_table$Ontology <- ontology

  # Convert p-value to numeric
  fishers_table$Fishers <- as.numeric(
    fishers_table$Fishers
  )

  # Multiple-testing correction
  fishers_table$FDR <- p.adjust(
    fishers_table$Fishers,
    method = "BH"
  )

  fishers_table
}
```

```{r}
# one big loop:

# ---------------------------------------------------------
# SETTINGS
# ---------------------------------------------------------

segments <- c("S1", "S2", "S3")

gene_sets <- c(
  "lox10_positive",
  "lox10_negative",
  "haci_positive",
  "haci_negative",
  "either_mutant_positive",
  "either_mutant_negative"
)

ontologies <- c("BP", "MF", "CC")


# ---------------------------------------------------------
# FUNCTION TO RUN ONE ANALYSIS
# ---------------------------------------------------------

run_one_GO_analysis <- function(segment, gene_set) {

  message(
    "Running: ",
    segment,
    " | ",
    gene_set
  )

  # Genes of interest
  selected_genes <- topGO_genes[[segment]][[gene_set]]

  # Create TopGO input vector
  myGenesF <- make_topGO_gene_vector(
    selected_genes,
    gene_names
  )

  # Run BP
  BP <- run_topGO(
    myGenesF,
    "BP",
    all_golist
  )

  # Run MF
  MF <- run_topGO(
    myGenesF,
    "MF",
    all_golist
  )

  # Run CC
  CC <- run_topGO(
    myGenesF,
    "CC",
    all_golist
  )

  # Combine
  combined <- rbind(BP, MF, CC)

  # Add metadata
  combined$Segment <- segment
  combined$GeneSet <- gene_set

  # Order by raw p-value
  combined <- combined[
    order(combined$Fishers),
  ]

  return(combined)
}
```

```{r}
# run everything:

# ---------------------------------------------------------
# RUN ALL ANALYSES
# ---------------------------------------------------------

GO_results <- list()

for (segment in segments) {

  for (gene_set in gene_sets) {

    result_name <- paste(
      segment,
      gene_set,
      sep = "_"
    )

    GO_results[[result_name]] <- run_one_GO_analysis(
      segment,
      gene_set
    )
  }
}
```

```{r}
# save the tables:

dir.create(
  "GO_results",
  showWarnings = FALSE
)

dir.create(
  "GO_results/tables",
  showWarnings = FALSE,
  recursive = TRUE
)

for (result_name in names(GO_results)) {

  write.csv(
    GO_results[[result_name]],
    file = paste0(
      "GO_results/tables/",
      result_name,
      ".csv"
    ),
    row.names = FALSE
  )
}
```



```{r}
# myGenes_BP.table <- run_topGO(myGenesF, "BP", all_golist)
# myGenes_MF.table <- run_topGO(myGenesF, "MF", all_golist)
# myGenes_CC.table <- run_topGO(myGenesF, "CC", all_golist)
```

combine results into one table:

```{r}
myGenesGO.table <- (rbind(myGenes_BP.table, myGenes_MF.table, myGenes_CC.table))
myGenesGO.table <- myGenesGO.table[order(myGenesGO.table$Fishers),]
```

MF (molecular function), BP (biological process), CC (cellular component)

```{r}
# write.csv(myGenesGO.table, "go_terms_s2.csv", row.names = F)
```

Plot it:

```{r}
myGenes_BP.table$Fishers <- as.numeric(myGenes_BP.table$Fishers)

myGenes_CC.table$Fishers <- as.numeric(myGenes_CC.table$Fishers)

myGenes_MF.table$Fishers <- as.numeric(myGenes_MF.table$Fishers)
```


Bubble plot:

```{r}
myBubblePlot <- function(table = myGenes_BP.table, term = "BP") {
  ggplot(table,
         aes(
           x = Term,
           y = -log10(Fishers),
           size = Significant,
           fill = -log10(Fishers)
         )) +
    expand_limits(y = 1) +
    geom_point(shape = 21) +
    #   scale_size(range = c(2.5, 12.5)) +
    scale_fill_continuous(low = 'royalblue', high = 'tomato') +
    xlab('') + ylab('-log10 (p value') +
    labs(title = paste0('GO Analysis (', term, ")")) +
    geom_hline(
      yintercept = c(-log10(0.05), -log10(0.01), -log10(0.001)),
      linetype = c("dotted", "dotted", "dotted"),
      colour = c("black", "black", "black"),
      size = c(1, 1, 1)
    ) +
    theme_bw(base_size = 12) +
    theme(
      legend.position = 'right',
      legend.background = element_rect(),
      plot.title = element_text(vjust = 1),
      plot.subtitle = element_text(vjust = 1),
      plot.caption = element_text(vjust = 1),
      
      axis.text.x = element_text(hjust = 1.10),
      axis.text.y = element_text(vjust = 0.5),
      axis.line = element_line(colour = 'black'),
      legend.key = element_blank(),
      legend.key.size = unit(1, "cm"),
    ) +
    coord_flip()
}
```

Barplot:
```{r}
myBarPlot <- function(table = myGenes_BP.table, term = "BP") {
  table <- table %>% mutate(word = case_when(Significant == 1 ~ "hit",
                                             TRUE ~ "hits"))
  ggplot(table, aes(reorder(Term, -log10(Fishers)),
                    -log10(Fishers), fill = Ontology)) +
    geom_bar(stat = 'identity') +
    scale_fill_manual(values = c(
      "BP" = "#9531b8",
      "MF" = "#9dd745",
      "CC" = "#da4218"
    )) +
    geom_text(aes(label = paste(Significant, word, "of", Annotated)),
              hjust = 2,
              color = "white") +
    theme_classic(base_size = 14) +
    ylab("- log10 p-value") +
    xlab("") +
    labs(title = paste0('GO Analysis (', term, ")")) +
    scale_y_continuous(expand = expansion(mult = c(0, .1)),
                       breaks = scales::pretty_breaks()) +
    theme(legend.position = "none") +
    coord_flip()
}
```

```{r}
plot_S1_bp1 <- myBubblePlot(table = myGenes_BP.table, term = "BP")
plot_S1_bp1

plot_S1_cc1 <- myBubblePlot(table = myGenes_CC.table, term = "CC")
plot_S1_cc1

plot_S1_mf1 <- myBubblePlot(table = myGenes_MF.table, term = "MF")
plot_S1_mf1
```

```{r}
plot_S1_bp2 <- myBarPlot(table = myGenes_BP.table, term = "BP")
plot_S1_bp2

plot_S1_cc2 <- myBarPlot(table = myGenes_CC.table, term = "CC")
plot_S1_cc2

plot_S1_mf2 <- myBarPlot(table = myGenes_MF.table, term = "MF")
plot_S1_mf2
```



## 5.2. KEGG


